---
title: "Prediction of 2020 American federal election using a regression
model"
author: "Hongrui Zhu"
date: "01/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(haven)
library(tidyverse)
library(pROC)
#### *****Data Selection & Cleaning*****####
#### Survey data -Load and select####
raw_data_survey <- read_dta("/Users/zhuhongrui/Desktop/ns20200625/ns20200625.dta")
raw_data_survey <- labelled::to_factor(raw_data_survey)

reduced_data_survey <- 
  raw_data_survey %>% 
  select(vote_2020,
         vote_intention,
         registration,
         age,
         gender,
         state,
         household_income,
         race_ethnicity)
#Adjust Data types
reduced_data_survey$age<-as.numeric(reduced_data_survey$age)
# Filter on survey data
#filter only on the people that are both registered & intented to vote 

#(Assuming people will vote unless they explicitly say no)
filtered_data_survey<-reduced_data_survey %>% 
                          filter(registration=="Registered"&
                                   vote_intention!="No, I am not eligible to vote"&
                                   vote_intention!="No, I will not vote but I am eligible"&
                                   (vote_2020=="Donald Trump"|vote_2020=="Joe Biden")
                          )
#Drop NAs
filtered_data_survey<-na.omit(filtered_data_survey)

rm(raw_data_survey,reduced_data_survey)

####Census data- Load and select####
raw_data_census <- read_dta("/Users/zhuhongrui/Desktop/ns20200625/usa_00003.dta.gz")
raw_data_census <- labelled::to_factor(raw_data_census)
# Just keep some variables
reduced_data_census <- 
  raw_data_census %>% 
  select(perwt,
         citizen,
         age,
         sex, 
         stateicp,
         inctot,
         race
         )
#Change data types
reduced_data_census$age<-as.numeric(reduced_data_census$age)
#reduced_data_census$inctot<-as.numeric(reduced_data_census$inctot)
#Filter Census data- only keeping those who can vote
filtered_data_census<-reduced_data_census %>% filter(age>=18 & (citizen=="naturalized citizen"|citizen=="born abroad of american parents"))

#Adjust some NAs
filtered_data_census$intot<-ifelse(filtered_data_census$inctot==9999999,
                                      NaN,filtered_data_census$inctot)


#Drop NAs (222298/228159, 97% data kept)
filtered_data_census<-na.omit(filtered_data_census)

rm(raw_data_census,reduced_data_census)


####Map data style between survey & census####

####Map sex/gender column####
unique(filtered_data_census$sex)
unique(filtered_data_survey$gender)
filtered_data_census$sex<-ifelse(filtered_data_census$sex=="female","Female","Male")

#rename census column so that variable names match
filtered_data_census<-rename(filtered_data_census,gender=sex)

unique(filtered_data_census$gender)
unique(filtered_data_survey$gender)


#####Create Age group in both datasets####
filtered_data_survey<-filtered_data_survey %>% 
                      mutate(agegroup = case_when(age <=20 ~ '20 or less',
                                             age >20  & age <= 35 ~ '21 to 35',
                                             age >35  & age <= 50 ~ '35 to 50',
                                             age >50  & age <= 65 ~ '50 to 65',
                                             age >65  & age <= 80 ~ '65 to 80',
                                             age >80 ~ 'above 80'
                                             )) 
filtered_data_census<-filtered_data_census %>% 
  mutate(agegroup = case_when(age <=20 ~ '20 or less',
                              age >20  & age <=  35 ~ '21 to 35',
                              age >35  & age <= 50 ~ '35 to 50',
                              age >50  & age <= 65 ~ '50 to 65',
                              age >65  & age <= 80 ~ '65 to 80',
                              age >80 ~ 'above 80' 
  )) 

unique(filtered_data_census$agegroup)
unique(filtered_data_survey$agegroup)

####Map Sate/Stateicp####
filtered_data_census<-filtered_data_census %>% 
  mutate(state = case_when(stateicp=="alabama"~"AL",
                           stateicp=="alaska"~"AK",
                           stateicp=="arizona"~"AZ",
                           stateicp=="arkansas"~"AR",
                           stateicp=="california"~"CA",
                           stateicp=="colorado"~"CO",
                           stateicp=="connecticut"~"CT",
                           stateicp=="delaware"~"DE",
                           stateicp=="florida"~"FL",
                           stateicp=="georgia"~"GA",
                           stateicp=="hawaii"~"HI",
                           stateicp=="idaho"~"ID",
                           stateicp=="illinois"~"IL",
                           stateicp=="indiana"~"IN",
                           stateicp=="iowa"~"IA",
                           stateicp=="kansas"~"KS",
                           stateicp=="kentucky"~"KY",
                           stateicp=="louisiana"~"LA",
                           stateicp=="maine"~"ME",
                           stateicp=="maryland"~"MD",
                           stateicp=="massachusetts"~"MA",
                           stateicp=="michigan"~"MI",
                           stateicp=="minnesota"~"MN",
                           stateicp=="mississippi"~"MS",
                           stateicp=="missouri"~"MO",
                           stateicp=="montana"~"MT",
                           stateicp=="nebraska"~"NE",
                           stateicp=="nevada"~"NV",
                           stateicp=="new hampshire"~"NH",
                           stateicp=="new jersey"~"NJ",
                           stateicp=="new mexico"~"NM",
                           stateicp=="new york"~"NY",
                           stateicp=="north carolina"~"NC",
                           stateicp=="north dakota"~"ND",
                           stateicp=="ohio"~"OH",
                           stateicp=="oklahoma"~"OK",
                           stateicp=="oregon"~"OR",
                           stateicp=="pennsylvania"~"PA",
                           stateicp=="rhode island"~"RI",
                           stateicp=="south carolina"~"SC",
                           stateicp=="south dakota"~"SD",
                           stateicp=="tennessee"~"TN",
                           stateicp=="texas"~"TX",
                           stateicp=="utah"~"UT",
                           stateicp=="vermont"~"VT",
                           stateicp=="virginia"~"VA",
                           stateicp=="washington"~"WA",
                           stateicp=="west virginia"~"WV",
                           stateicp=="wisconsin"~"WI",
                           stateicp=="wyoming"~"WY",
                           stateicp=="district of columbia"~"DC")) 
 filtered_data_census$stateicp<-NULL

unique(filtered_data_census$state)
unique(filtered_data_survey$state)
 
####Map income####
x<-unique(filtered_data_survey$household_income)
min(filtered_data_census$intot)
max(filtered_data_census$intot)
filtered_data_census<-filtered_data_census %>% 
  mutate(household_income = case_when(intot<=14999 ~ "Less than $14,999",
                                      intot>=15000 & intot<=19999~"$15,000 to $19,999",
                                      intot>=20000 & intot<=24999~"$20,000 to $24,999",
                                      intot>=25000 & intot<=29999~"$25,000 to $29,999",
                                      intot>=30000 & intot<=34999~"$30,000 to $34,999",
                                      intot>=35000 & intot<=39999~"$35,000 to $39,999",
                                      intot>=40000 & intot<=44999~"$40,000 to $44,999",
                                      intot>=45000 & intot<=49999~"$45,000 to $49,999",
                                      intot>=50000 & intot<=54999~"$50,000 to $54,999",
                                      intot>=55000 & intot<=59999~"$55,000 to $59,999",
                                      intot>=60000 & intot<=64999~"$60,000 to $64,999",
                                      intot>=65000 & intot<=69999~"$65,000 to $69,999",
                                      intot>=70000 & intot<=74999~"$70,000 to $74,999",
                                      intot>=75000 & intot<=79999~"$75,000 to $79,999",
                                      intot>=80000 & intot<=84999~"$80,000 to $84,999",
                                      intot>=85000 & intot<=89999~"$85,000 to $89,999",
                                      intot>=90000 & intot<=94999~"$90,000 to $94,999",
                                      intot>=95000 & intot<=99999~"$95,000 to $99,999",
                                      intot>=100000 & intot<=124999~"$100,000 to $124,999",
                                      intot>=125000 & intot<=149999~"$125,000 to $149,999",
                                      intot>=150000 & intot<=174999~"$150,000 to $174,999",
                                      intot>=175000 & intot<=199999~"$175,000 to $199,999",
                                      intot>=200000 & intot<=249999~"$200,000 to $249,999",
                                      intot>=250000~"$250,000 and above"
                                      )) 

filtered_data_census$intot<-NULL

unique(filtered_data_census$household_income)
unique(filtered_data_survey$household_income)

####Map race####
length(unique(filtered_data_survey$race_ethnicity))
length(unique(filtered_data_census$race))

otherasian<-c("Asian (Asian Indian)","Asian (Vietnamese)","Asian (Other)","Asian (Korean)","Asian (Filipino)",
              "Pacific Islander (Native Hawaiian)","Pacific Islander (Other)",
              "Pacific Islander (Samoan)","Pacific Islander (Guamanian)")
#survey data
filtered_data_survey<-filtered_data_survey %>% 
  mutate(race = case_when(race_ethnicity =="Asian (Japanese)" ~ 'Japanese',
                          race_ethnicity =="Asian (Chinese)" ~ 'Chinese',
                          race_ethnicity %in% otherasian ~"other asian or pacific islander",
                          race_ethnicity =="White" ~ 'White',
                          race_ethnicity =="Black, or African American" ~ 'Black, or African American',
                          race_ethnicity =="Some other race" ~ 'Other race',
                          race_ethnicity=="American Indian or Alaska Native"~"American Indian or Alaska Native",
                          race_ethnicity=="Other race "~"Other race"
  )) 
filtered_data_survey$race_ethnicity<-NULL

#census data
filtered_data_census<-filtered_data_census %>% 
  mutate(race2 = case_when(race=="white"~"White",
                           race=="chinese"~"Chinese",
                           race=="black/african american/negro"~"Black, or African American",
                           race=="two major races"~"Other race",
                           race=="other race, nec"~"Other race",
                           race=="japanese"~"Japanese",
                           race=="american indian or alaska native"~"American Indian or Alaska Native",
                           race=="three or more major races"~"Other race",
                           race=="other asian or pacific islander"~"other asian or pacific islander"
  )) 
unique(filtered_data_census$race2)

filtered_data_census$race<-filtered_data_census$race2
filtered_data_census$race2<-NULL

unique(filtered_data_census$race)
unique(filtered_data_survey$race)


####Populate Datset for modelling####
filtered_data_survey%>% select(vote_2020,age,agegroup,gender,state,household_income,race)->survey.data
filtered_data_census%>% select(perwt,age,agegroup,gender,state,household_income,race)->census.data


####Create cell variable in both datasets####
#create cells- Do not create too many cells as the sample obs number is limited
survey.data$cell<-paste(survey.data$household_income,survey.data$race)
census.data$cell<-paste(census.data$inctot,census.data$race)


####Convert variables to factors#####
f.cols.survey<-c("agegroup","gender","state","household_income" ,"race", "cell","vote_2020")
survey.data[f.cols.survey] <- lapply(survey.data[f.cols.survey], factor) 
f.cols.census<-c("agegroup","gender","state","household_income" ,"race", "cell","perwt")
census.data[f.cols.census] <- lapply(census.data[f.cols.census], factor) 
survey.data$vote_2020 <- relevel(survey.data$vote_2020, ref = "Donald Trump")    #To predict probability of voting for Biden (Trump as ref)

####Count number of cells####
length(unique(survey.data$cell))  
length(unique(census.data$cell))



####*****Multi-level regression*****####
library(lme4)
library(brms)
library(tidybayes)
library(caret)
library(ROCR)

####Model 1####
model_logit1 <- glmer(vote_2020~(1+race|cell)+gender+state+household_income,
                     data = survey.data, 
                     family=binomial)

summary(model_logit1)

prob.1<-predict(model_logit1,type=c('response'))
result_model1<-ifelse(prob.1>=0.5,"Joe Biden","Donald Trump")
survey.data.result<-cbind(survey.data,result_model1)

#Logistic: ROC Curve 
roc.1 <- roc(survey.data.result$vote_2020, prob.1)
auc(roc.1)
plot(roc.1, auc.polygon=TRUE, print.auc = TRUE,asp = NA)




####*****Post-Stratification*****####

####Apply model on census data####
vote_2020_prob<-predict(model_logit1,census.data[,c("gender","state",
                                    "household_income","race","cell")],type="response",allow.new.levels = TRUE)
vote_2020_pred<-ifelse(vote_2020_prob>0.5,"Joe Biden","Donald Trump")
census.data.result<-cbind(census.data,vote_2020_pred)

####calculate total votes based on person weight####
census.data.result$trump_votes<-ifelse(census.data.result$vote_2020_pred=="Donald Trump",census.data.result$perwt,0)
census.data.result$biden_votes<-ifelse(census.data.result$vote_2020_pred=="Joe Biden",census.data.result$perwt,0)

####Calculate vote per state####
census.data.result %>% group_by(state) %>% summarise(Trump=sum(trump_votes),Biden=sum(biden_votes))->predicted_states
predicted_states$winner<-ifelse(predicted_states$Trump>predicted_states$Biden,
                                "Donald Trump","Joe Biden")

####Mutate electoral votes per state#### (from google)
predicted_states<-predicted_states %>% 
  mutate(electoral_votes = case_when(state=="CA"~55,state=="TX"~38,state=="FL"~29,state=="NY"~29,state=="IL"~20,state=="PA"~20,state=="OH"~18,
                                     state=="GA"~16,state=="MI"~16,state=="NC"~15,state=="NJ"~14,state=="VA"~13,state=="WA"~12,state=="AZ"~11,
                                     state=="IN"~11,state=="MA"~11,state=="TN"~11,state=="MD"~10,state=="MN"~10,state=="MO"~10,state=="WI"~10,
                                     state=="AL"~9,state=="CO"~9,state=="SC"~9,state=="KY"~8,state=="LA"~8,state=="CT"~7,state=="OK"~7,
                                     state=="OR"~7,state=="AR"~6,state=="IA"~6,state=="KS"~6,state=="MS"~6,state=="NV"~6,state=="UT"~6,
                                     state=="NE"~5,state=="NM"~5,state=="WV"~5,state=="HI"~4,state=="ID"~4,state=="ME"~4,state=="NH"~4,
                                     state=="RI"~4,state=="AK"~3,state=="DE"~3,state=="MT"~3,state=="ND"~3,state=="SD"~3,state=="VT"~3,
                                     state=="WY"~3,state=="DC"~3
  )) 


predicted_states %>% group_by(winner) %>% summarise(total_votes=sum(electoral_votes))->election_result
election_result
```

##Model##

In this project, I choose to use generalized mixed effect model with binomial family. I choose this model is because it contains both random effects and fixed effects, and it is useful because the data I have here has more than one source of variables, by using this model, repeated measurement of the same observation would happen, and that would give us more information than other models. In this model, we have fixed effect gender, state and household income, telling us how different gender, state and income would affect their willness to vote, for example, genderm tells that male are less likely to vote for Trump.

In the model we built, there are several variables. The variable gender tells us whether the respondents are male or female, and the variable state tells us, by providing two-letters, which state the respondent is from. The variable household_income, provides information about how much money the respondent's family makes each year. And the last variable in the model is the random effect of race, telling people what intention that different races are going to vote in this election. 

In addition, I used post-stratification, it is a method to adjust sample weights so we can get the account of the underrepresented group in the population. This is a necessary step because the data we are dealing with contains too much data across the American, people are various, for example, white and black are the majority of the population, and others are the minority. However, in this project, I need to consider everyone's opinion, not just the majority's, so the post-stratification is necessary here.

##Result##

```{r}
plot(main = "Figure 1.1 ROC plot", roc.1, auc.polygon=TRUE, print.auc = TRUE,asp = NA)
```


As we can see above(Figure 1.1), the AUC of the model I built is 0.702, suggesting that it is an acceptable model, 0.702 AUC means that there is about 70% chance that the model can predict correctly. Most importantly, the result we initially looking for, who would win the election, the model concludes that Biden would get 303 tickets and Trump would get 235 tickets so Biden would eventually win the election. We estimate that the proportion of voters in favor of voting for Trump to be 0.43, suggesting that based on the model we built, more people are willing to vote for Biden, and Biden would eventually win the election. For coefficient we have for this model, the coefficient of "gendermale" is -0.4496, means that male are less likely to vote for Trump compared to female. And for income variable, family income that is between 200,000 to 249,999 is more willing to vote for Biden, and p-value of it is less than 0.05, suggesting that it is significant.  

##Discussion##


The most time I spent on this assignment is the preparation, cleaning data, and mapping two datasets that spent a lot of time. I removed the NA variables, and make the variables in both datasets match each other, so we can compare different datasets. After finished preparation, the work was just to build a model, do the post-stratification, and make the prediction. Basing on the result we have, the model suggests that Biden would get more votes and win the election of 2020, and females are significantly intending to vote for Biden more.

For the weakness of the model, the biggest problem of the model is that I am supposed to make two models and compare which one is better, but the dataset we are dealing with is just too huge, it took a lot of time to run the model and get the result, so I was not able to make a different model. The huge dataset makes the analysis difficult, my computer stop working many times when I was doing the process, and that forced me to reduce the dataset and simplify the model, at the end, only a limited number of data was processed by the model and the model was, in my opinion, too simple for the problem we are looking for. However, this is no way for me to solve this problem unless I can get a more powerful device to compute. Another weakness is that the data was collected in June, and it has been a few months ago, during the last few months, America is experiencing a terrible situation due to COVID-19, and this might have a huge effect on the result of the 2020 election vote, but I am not able to put that into consideration since no such data is provided.

For the next step, if I can do the analysis one more time, I would try to put more observations into consideration, so more observations that are in the minority groups can be analyzed. The goal of this model is to predict the 2020 election vote not just basing on the majority, but everyone who would vote for the election, so more situations are considered, the more precise the prediction would be.



##Reference##

-Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software,
  4(43), 1686, https://doi.org/10.21105/joss.01686
  
-Xavier Robin, Natacha Turck, Alexandre Hainard, Natalia Tiberti, Frédérique Lisacek,
  Jean-Charles Sanchez and Markus Müller (2011). pROC: an open-source package for R and
  S+ to analyze and compare ROC curves. BMC Bioinformatics, 12, p. 77.  DOI:
  10.1186/1471-2105-12-77 <http://www.biomedcentral.com/1471-2105/12/77/>

-Hadley Wickham and Evan Miller (2020). haven: Import and Export 'SPSS', 'Stata' and
  'SAS' Files. R package version 2.3.1. https://CRAN.R-project.org/package=haven
  
-Sing T, Sander O, Beerenwinkel N, Lengauer T (2005). “ROCR: visualizing classifier
performance in R.” _Bioinformatics_, *21*(20), 7881. <URL:
http://rocr.bioinf.mpi-sb.mpg.de>.

-Max Kuhn (2020). caret: Classification and Regression Training. R package version
  6.0-86. https://CRAN.R-project.org/package=caret
  
-Paul-Christian Bürkner (2017). brms: An R Package for Bayesian Multilevel Models Using
  Stan. Journal of Statistical Software, 80(1), 1-28. doi:10.18637/jss.v080.i01

Paul-Christian Bürkner (2018). Advanced Bayesian Multilevel Modeling with the R Package
  brms. The R Journal, 10(1), 395-411. doi:10.32614/RJ-2018-017

-Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear
  Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48.
  doi:10.18637/jss.v067.i01.